A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-02-16 paper/paper.tex
Minor
Add renaming properties of the logical relation

The logical relation we define in Figure~\ref{fig:logrel} does not
include cases for variables. Likewise, for the base types such as
boolean and equality, the output set $[[S]]$ contains only terms that
evaluate to closed terms. To generalize our logical relation to open
terms, we define the semantic typing judgment by closing the open
terms with a substitution whose codomain consists of terms that
respect the interpretation of the types from the context. The full
definitions of well-formed substitution ($[[rho |= G]]$), semantic
typing ($[[ G |= a : A]]$), and semantic context well-formedness
($[[|= G]]$) are presented in Figure~\ref{fig:semtyping}.

@@

Next, we show some non-trivial cases of the fundamental theorem as
top-level lemmas and leave the remaining cases as exercises for the reader.
\begin{lemma}[ST-Var]
  \label{lemma:stvar}
  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= var i : G i < up Suc i  >]]$.
\end{lemma}
\begin{proof}
  TODO: formulate renaming
\end{proof}

>>>>

The logical relation we define in Figure~\ref{fig:logrel} does not
include cases for variables. Likewise, for the base types such as
boolean and equality, the output set $[[S]]$ contains only terms that
evaluate to closed terms. To generalize our logical relation to open
terms, we define the semantic typing judgment by closing the open
terms with a substitution whose codomain consists of terms that
respect the interpretation of the types from the context. The full
definitions of well-formed substitution ($[[rho |= G]]$), semantic
typing ($[[ G |= a : A]]$), and semantic context well-formedness
($[[|= G]]$) are presented in Figure~\ref{fig:semtyping}. In the
definition of $[[|= G]]$, we use the notation $[[drop i G]]$ to denote
the typing context obtained by dropping the last $[[i]]$ elements of
$[[G]]$. When $[[G]]$ has less than $[[i]]$ elements, $[[drop i G]]$
returns the empty list.

@@

Next, we show some non-trivial cases of the fundamental theorem as
top-level lemmas and leave the remaining cases as exercises for the
reader.

First, we formulate the definition of valid renamings and prove that
semantic typing satisfies renaming. We say that $[[xi]]$ is
valid from the context $[[G]]$ to the context $[[D]]$ if the
following condition holds.
\[ \forall i \text{, if } i < | [[G]] |\text{, then }[[ ren xi i < |D| ]] \text{ and }
  [[D ren xi
  i < up Suc ren xi i > = G i < up Suc i > < xi >]] \]

\begin{lemma}[Truncate is valid]
  \label{lemma:validtruncate}
  If $[[i < |G|]]$, then $[[up i]]$ is a valid renaming from $[[drop i
  G]]$ to $[[G]]$.
\end{lemma}
\begin{proof}
  By the definition of a valid renaming, we must show that given $[[j
  < |drop i G|]] = [[|G|]] - [[i]]$, then the following conditions hold:
  \begin{itemize}
  \item $[[ren up i j]] = i + j  < [[| G | ]]$
  \item $[[G ren up i j < up Suc ren up i j > = drop i G j < up Suc j > < up i >]]$
  \end{itemize}
  The first bullet point is immediate from the fact that $[[j]] <
  [[|G|]] - [[i]]$. The second bullet follows from unfolding the
  definitions, and the fact that $[[drop i G j]] = [[G]](i + j)$ when
  $[[i]] + [[j]] < [[|G|]]$.
\end{proof}

\begin{lemma}[Renaming for $[[rho |= G]]$]
  \label{lemma:renamingval}
  If $[[rho |= D]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
  $[[D]]$, then we have $[[rho0 |= G]]$ where $[[rho0 i = rho ren xi i]]$.
\end{lemma}
\begin{proof}
  Unfolding the definition of $[[rho0 |= G]]$, the goal is to show
  that for all $i,j,$ and $S$, if $[[i < |G|]]$ and $[[InterpR j (G i
  < up Suc i > ) { rho0 } S ]] \text{, then } [[rho0 i]] = [[rho ren xi i in S]]$.

  By the definition of $[[rho0]]$ and the validity of $[[xi]]$, we
have $[[G i < up Suc i > { rho0 }]] = [[G i < up Suc i > < xi > { rho
} ]] = [[D ren xi i < up Suc ren xi i > { rho }]]$. From $[[rho |=
D]]$ and $[[ren xi i < | D |]]$, we have $[[InterpR j D ren xi i < up Suc ren xi i > { rho }
S]]$ and $[[rho ren xi i in S]]$. Done.
\end{proof}

\begin{lemma}[Renaming for $[[G |= a : A]]$]
  \label{lemma:semrenaming}
  If $[[G |= a : A]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
  $[[D]]$, then $[[D |= a < xi > : A < xi > ]]$.
\end{lemma}
\begin{proof}
  Immediate from the definition of semantic typing and Lemma~\ref{lemma:renamingval}.
\end{proof}

\begin{lemma}[ST-Var]
  \label{lemma:stvar}
  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= var i : G i < up Suc i  >]]$.
\end{lemma}
\begin{proof}
  Suppose $[[rho |= G]]$. By the definition of semantic typing, we
need to show that there exists some $[[j]]$ and $[[S]]$ such that
  \begin{itemize}
  \item $[[InterpR j G i < up Suc i > { rho } S]]$
  \item $[[rho i in S]]$
  \end{itemize}
  From the definition of $[[|= G]]$, we know that there exists some
  $[[j]]$ such that $[[drop Suc i G |= G i : Set j]]$. By
  Lemma~\ref{lemma:validtruncate}, we know that $[[up Suc i]]$ is a
  valid renaming from $[[drop Suc i G]]$ to $[[G]]$. By
  Lemma~\ref{lemma:semrenaming}, we deduce $[[G |= G i < up Suc
  i > : Set j]]$. By the identity from Lemma~\ref{lemma:setinv}, we
  know that $[[InterpR j G i < up Suc i > { rho } S]]$. It now
  suffices to show $[[rho i in S]]$, but that is immediate from the
  definition of $[[rho |= G]]$.
\end{proof}
